Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S.

### 2023-04-24 Draft.Rmd
removed pca

---
title: "Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S."
author: "Aileen Li, Nyah Strickland"
date: "2023-04-18"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

# Calculate principal components on scaled training data
pca_result <- train_scaled %>% dplyr::select(where(is.numeric)) %>%
  prcomp()

# Plot scree plot
pca_result %>% fviz_eig()

# Plot table to show explained variance % by PCs
summary(pca_result)

# Threshold of Explained Variance = 80%
# Calculated the sum of PCs with highest explained variance until threshold was meant
0.3253 + 0.09749 + 0.07984 + 0.06501 + 0.05196 + 0.04147 + 0.03445 +
0.02926 + 0.02563 + 0.0220 + 0.02127 + 0.0188

PC1 <- pca_result$rotation[,1]
PC1_scores <- abs(PC1)
PC1_scores_ordered <- sort(PC1_scores, decreasing = TRUE)
names(PC1_scores_ordered)

# Reduce dimension of dataset
trunc <- pca_result$x[,1:12] %*% t(pca_result$rotation[,1:12])
# Add the center (and re-scale) back to data 
if(all(pca_result$scale != FALSE)){
  trunc <- scale(trunc, center = FALSE , scale=1/pca_result$scale) 
  } 
if(all(pca_result$center != FALSE)){ 
  trunc <- scale(trunc, center = -1 * pca_result$center, scale=FALSE) 
  } 
dim(trunc); dim(train_scaled)

# unhelpful plot -- delete
# plot(pca_result$x[,1], pca_result$x[,2])

# Show which features have the most influence
as_tibble(pca_result$rotation, 
          rownames = "variable") %>% 
    ggplot(aes(variable, PC1)) +
    geom_point() +
    coord_flip()

as_tibble(pca_result$rotation, 
          rownames = "variable") %>% 
    ggplot(aes(variable, PC2)) +
    geom_point() +
    coord_flip()

# Plot clusters between PC1 and PC2 using a biplot
fviz_pca_var(pca_result)

# Reduce dimension of dataset
trunc <- pca_result$x[,1:12] %*% t(pca_result$rotation[,1:12])
# Add the center (and re-scale) back to data 
if(all(pca_result$scale != FALSE)){
  trunc <- scale(trunc, center = FALSE , scale=1/pca_result$scale) 
  } 
if(all(pca_result$center != FALSE)){ 
  trunc <- scale(trunc, center = -1 * pca_result$center, scale=FALSE) 
  } 
dim(trunc); dim(train_scaled)

@@

@@

The most challenging aspect of this project was figuring out the process to determine which predictors were the best to use for our models. We ran into many debugging issues and Googled a lot to resolve our uncertainties. We learned that finding the best predictors for a model can be a lengthy process and that running a new model (e.g. preprocessing steps for XG Boost) often may require lots of research into what is needed for the model to run before it can actually run.

@@

**Include acknowledgments for any help received. If a group project, report the contribution of each member (i.e. who did what?).**

## Acknowledgements
**Aileen Li & Nyah Strickland:**
- Intro

>>>>

---
title: "Predicting Ambient Air Pollution (PM2.5) Across the Contiguous U.S."
author: "Aileen Li, Nyah Strickland"
date: "2023-04-18"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

@@

@@

We learned that we overfit our Xgboost model to the data, which is why the prediction performance was poor for the testing dataset. This can be seen in our model having an RMSE value of 1.38 despite the RMSE of the model fit to the training data being 0.23. In general, we learned that understanding the disadvantages and advantages of each model is helpful toward understanding why a model fails or succeeds at the training and testing stages. For instance, we learned the Xgboost is helpful for unbalanced data and that Random Forest Regression performs well for datasets with a lot of noise.

@@

The most challenging aspect of this project was figuring out the process to determine which predictors were the best to use for our models. We ran into many debugging issues and Googled a lot to resolve our uncertainties. We learned that finding the best predictors for a model can be a lengthy process and that running a new model (e.g. preprocessing and hyperparameter tuning steps for XG Boost and Random Forest Regression) often may require lots of research into what is needed for the model to run before it can actually run.

@@

**Include acknowledgments for any help received. If a group project, report the contribution of each member (i.e. who did what?).**
## Acknowledgements
**Aileen Li & Nyah Strickland:**
- Intro
